---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```


<!-- badges: start -->
  [![rOG-badge](https://ropengov.github.io/rogtemplate/reference/figures/ropengov-badge.svg)](https://ropengov.org/)
  [![R build status](https://github.com/rOpenGov/geofi//workflows/R-CMD-check/badge.svg)](https://github.com/rOpenGov/geofi/actions/)
  [![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/)
  [![codecov](https://codecov.io/gh/rOpenGov/geofi/branch/master/graph/badge.svg?token=yJSHHMSSKs)](https://app.codecov.io/gh/rOpenGov/kelaopendata)
[![CRAN published](https://www.r-pkg.org/badges/version/kelatools)](https://www.r-pkg.org/pkg/kelatools)
<!-- badges: end -->


# kelaopendata - Access open data from National social insurance institution of Finland

Finnish national social insurance institution [Kela](https://www.kela.fi/) publishes open data on recipients, reimbursements and costs of various social security benefits in Finland at <https://www.avoindata.fi/data/fi/organization/kela>. `kelaopendata`-package takes advantage of modern big data technologies such as [duckdb](https://duckdb.org/) and [Apache Parquet](https://parquet.apache.org/) and provides fast and straightforward access to data. 

Below is a one example on how you can access data using `kelaopendata`. Please have a closer look at the [vignettes](https://ropengov.github.io/kelaopendata/articles/index.html) for more comprehensive use cases. 


## Installation and use

```{r eval = FALSE}
# Not yet published in CRAN

# Install development version from GitHub
remotes::install_github("rOpenGov/kelaopendata")
```


## List available datasets

```{r}
library(kelaopendata)
dsets <- list_datasets()
print(dsets, n = 50)
```

Print first 10 rows of most recently updated data that is `r kelaopendata::get_metadata(data_id = dsets[1,]$id)$title`

```{r}
kelaopendata::get_data(data_id = dsets[1,]$id, sql = "LIMIT 10")
```

Also, there is a web application at <https://kelaopendata.vaphana.com/> that can be used to browse this data resource.


## Disclaimer

This package is in no way officially related to or endorsed by KELA.


