---
title: "Fetching data using kelaopendata"
author: "Markus Kainu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geocoding with R and geofi}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, 
  warning = FALSE,
  fig.height = 7,  
  fig.width = 7,
  dpi = 75
)
```


**Installation**

`kelaopendata` can be installed from Github using

```{r, eval = FALSE}
# Install development version from GitHub
remotes::install_github("ropengov/kelaopendata")
```

```{r}
# Let's first create a function that checks if the suggested 
# packages are available
check_namespaces <- function(pkgs){
  return(all(unlist(sapply(pkgs, requireNamespace,quietly = TRUE))))
}
```

## List available datasets

```{r}
library(kelaopendata)
library(dplyr)

dsets <- list_datasets()
print(dsets, n = 50)
```

## Obtaining data on Financial aid for students (opintotuki)

### Metadata

For this example we choose "Financial aid for students" as our benefit of interest.  First we download the metadata and print the description field

```{r}
d_id <- dsets[dsets$name == "opintotuen-saajat-ja-maksetut-tuet", ]$id
meta <- get_metadata(data_id = d_id)
meta$description
```

And then a more technical overview of the content of data set, containing names of the csv-files, csv-dialect and values and types of each indicator in the data. 

```{r}
jsonlite::toJSON(meta$resources, pretty = T)
```

A more dense view of variables and their types and descriptions can be printed with

```{r}
meta$resources$schema$fields[[1]] |>
  select(-values) |>
  as_tibble()
```




### Quering, downloading and plotting the data

Let's query data on recipients of Student loan in the city of Turku using `kelaopendata::get_data()`-function

```{r}
d_opintotuki <- kelaopendata::get_data(
  data_id = d_id,
  sql = "WHERE etuus = 'Opintolainan valtiontakaus' AND
                            aikatyyppi = 'Vuosi' AND
                            kunta_nimi = 'Turku' AND
                            etuus = 'Opintolainan valtiontakaus' AND
                            oppilaitos_peruste = 'Viimeisin oppilaitos'
                           "
)
d_opintotuki
```

Next, let's filter the data locally in R a bit more.

```{r}
d_plot <- d_opintotuki %>%
  # Exclude
  filter(sukupuoli != "Tuntematon",!oppilaitosaste %in% c("Tieto puuttuu", "Yhteensä")) %>%
  mutate(oppilaitosaste = factor(
    oppilaitosaste,
    levels = c(
      "Yliopistot",
      "Ammattikorkeakoulut",
      "Ammatilliset oppilaitokset",
      "Lukiot",
      "Muut oppilaitokset",
      "Ulkomaiset oppilaitokset"
    )
  ))
```


Finally, let's draw a plot on recipients by gender and type of institution


```{r, fig.width=8, fig.height=12}
library(ggplot2)
ggplot(d_plot, aes(x = vuosi, y = saaja_lkm, fill = ikaryhma)) +
  geom_col(position = position_stack()) +
  facet_grid(oppilaitosaste ~ sukupuoli) +
  labs(title = "Recipients of government guarantee for a student loan in\nthe city of Turku in 2004 to 2024 by gender and type of institution") +
  theme_light()
```




```{r}
d_opintotuki <- kelaopendata::get_data(
  data_id = d_id,
  sql = "WHERE etuus = 'Opintolainan valtiontakaus' AND
                            aikatyyppi = 'Vuosi' AND
                            vuosi = '2023' AND
                            etuus = 'Opintolainan valtiontakaus' AND
                            oppilaitos_peruste = 'Viimeisin oppilaitos'
                           "
)
d_opintotuki

library(geofi)
muni <- get_municipalities()
```

Next, let's filter the data locally in R a bit more.

```{r}
d_plot <- d_opintotuki %>%
  # Exclude
  filter(sukupuoli != "Tuntematon",!oppilaitosaste %in% c("Tieto puuttuu", "Yhteensä")) %>%
  filter(oppilaitosaste == "Yliopistot",
         !is.na(sukupuoli)) %>% 
  group_by(kunta_nro,sukupuoli) %>% 
  summarise(saaja_lkm = sum(saaja_lkm)) %>% 
  ungroup() %>% 
  mutate(municipality_code = as.integer(kunta_nro))
d_plot_sf <- left_join(muni,d_plot)
```


Finally, let's draw a plot on recipients by gender and type of institution


```{r, fig.width=8, fig.height=12}
library(ggplot2)
ggplot(d_plot_sf, aes(fill = saaja_lkm)) +
  geom_sf() +
  facet_wrap(~ sukupuoli) +
  theme_light()
```

